---
title: "HW2-explore and wrangle"
date: 2016-10-05
output: html_document
---



## Setup

```{r setup, message = FALSE, results = 'hide'}
library(devtools)
library(rcfss)
library(tidyverse)
library(stringr)
library(scales)
```



## Part 1: Exploring clean data
Using your knowledge of dplyr and ggplot2, use summary statistics and graphs to answer the following questions:
1. In what month do the most gun deaths occur?
2. What is the most common intent in gun deaths? Do most people killed by guns die in suicides, homicides, or accidental shootings?
3. What is the average age of females killed by guns?
4. How many white males with at least a high school education were killed by guns in 2012?
5. Which season of the year has the most gun deaths? Assume that 
Winter = January-March
Spring = April-June
Summer = July-September
Fall = October-December
6. What is the relationship between race and intent? For example, are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r part 1}

data("gun_deaths")

reorder_bysize <- function(x) {
  factor(x, levels = names(sort(table(x), decreasing = TRUE)))
}

#1
gdeaths_tidy <- gun_deaths %>%
  mutate(month = as.factor(month.abb[month]))

ggplot(gdeaths_tidy, aes(reorder_bysize(month))) +
  geom_bar()

#2
gdeaths_tidy <- gun_deaths %>%
  mutate(intent = as.factor(intent))

ggplot(gdeaths_tidy[!is.na(gdeaths_tidy$intent), ], aes(reorder_bysize(intent))) +
  geom_bar()

#3
gdeaths_tidy <- gun_deaths %>%
  group_by(sex)

ggplot(gdeaths_tidy, aes(sex, age)) +
  geom_boxplot()

summarize(gdeaths_tidy, avgAge = mean(age, na.rm = TRUE))

#4
gdeaths_tidy <- gun_deaths %>%
  filter(education != "Less than HS", year == 2012, sex == "M", race == "White")

summarize(gdeaths_tidy, n = n())

#the answer from prof is without the criterion of white

#5
gdeaths_tidy <- gun_deaths %>%
  mutate(season = cut(month, breaks = c(0, 3, 6, 9, 12), labels = c("Winter", "Spring", "Summer", "Fall")))

# gdeaths_tidy$season[gdeaths_tidy$season %in% c(1, 2, 3)] <- "winter"
# gdeaths_tidy$season[gdeaths_tidy$season %in% c(4, 5, 6)] <- "Spring"
# gdeaths_tidy$season[gdeaths_tidy$season %in% c(7, 8, 9)] <- "Summer"
# gdeaths_tidy$season[gdeaths_tidy$season %in% c(10, 11, 12)] <- "Fall"

ggplot(gdeaths_tidy, aes(reorder_bysize(season))) +
  geom_bar()

#6
gdeaths_tidy <- gun_deaths

ggplot(gdeaths_tidy, aes(race, fill = intent)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = percent_format())

```



## Part 2: Tidying messy data
Tidy this data frame so that it adheres to the tidy data principles:

```{r part 2}

data("dadmom")
dad <- dadmom %>%
  transmute(famid, name = named, inc = incd, type = "dad")
mom <- dadmom %>%
  transmute(famid, name = namem, inc = incm, type = "mom")

dadmom_tidy <- bind_rows(dad, mom)
dadmom_tidy

```



## Part 3: Wrangling and exploring messy(ish) data
There are several ID variables which are useful for other types of research: for our purposes, the only ID variable you need to concern yourself with is caseIssuesId. Variables you will want to familiarize yourself with include: term, justice, justiceName, decisionDirection, majVotes, minVotes, majority, and chief. Pay careful attention in the documentation to how these variables are coded.

Once joined, use your data wrangling and exploratory data analysis skills to answer the following questions:
1. What percentage of cases in each term are decided by a one-vote margin (i.e. 5-4, 4-3, etc.)
2. In each term he served on the Court, in what percentage of cases was Justice Antonin Scalia in the majority? 
3A. Advanced challenge: Create a graph similar to above that compares the percentage for all cases versus non-unanimous cases (i.e. there was at least one dissenting vote)
4. In each term, what percentage of cases were decided in the conservative direction? 
5A. Advanced challenge: The Chief Justice is frequently seen as capable of influencing the ideological direction of the Court. Create a graph similar to the one above that also incorporates information on who was the Chief Justice during the term.

Terms run from October through June, so the 2015 term contains cases decided from October 2015 - June 2016

```{r part 3}

jcitation_legacy <- read_csv("data/SCDB_Legacy_03_justiceCentered_Citation.csv") %>%
  mutate(docket = as.integer(docket), adminAction = as.integer(adminAction), adminActionState = as.integer(adminActionState))
jcitation_modern <- read_csv("data/SCDB_2016_01_justiceCentered_Citation.csv")

jcitation_all <- bind_rows (jcitation_legacy, jcitation_modern) %>%
  transmute(caseIssuesId, term, justice, justiceName, decisionDirection, majVotes, minVotes, majority, chief)

jcitation_all

#1
jcitation_ref <- jcitation_all %>%
  distinct(caseIssuesId, .keep_all = TRUE) %>%
  group_by(term) %>%
  summarize(totalNumber = n())

jcitation_ref

jcitation_tidy <- jcitation_all %>%
  distinct(caseIssuesId, .keep_all = TRUE) %>%
  mutate(difference = majVotes - minVotes) %>%
  group_by(term, difference) %>%
  summarize(number = n()) %>%
  filter(difference == 1) %>%
  left_join(jcitation_ref, by = "term") %>%
  mutate(percentage = number / totalNumber)

jcitation_tidy

ggplot(jcitation_tidy, aes(term, percentage)) +
  geom_line() +
  scale_y_continuous(labels = percent_format())

#2
jcitation_ref <- jcitation_all %>%
  filter(justiceName == "AScalia") %>%
  group_by(term) %>%
  summarize(totalNumber = n())

jcitation_tidy <- jcitation_all %>%
  filter(justiceName == "AScalia", majority == 2) %>%
  group_by(term) %>%
  summarize(number = n()) %>%
  left_join(jcitation_ref, by = "term") %>%
  mutate(percentage = number / totalNumber)

jcitation_tidy

ggplot(jcitation_tidy, aes(term, percentage)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1), labels = percent_format())

#3A
jcitation_ref <- jcitation_all %>%
  filter(justiceName == "AScalia") %>%
  mutate(ifUnanimous = minVotes)

jcitation_ref$ifUnanimous[jcitation_ref$ifUnanimous > 0] <- "nonUnanimous"
jcitation_ref$ifUnanimous[jcitation_ref$ifUnanimous == 0] <- "unanimous"

jcitation_ref <- jcitation_ref %>%
  group_by(term, ifUnanimous) %>%
  summarize(totalNumber = n()) %>%
  spread(key = ifUnanimous, value = totalNumber) %>%
  mutate(totalNumber = nonUnanimous + unanimous)

jcitation_ref

jcitation_tidy <- jcitation_all %>%
  filter(justiceName == "AScalia", majority == 2) %>%
  mutate(ifUnanimous = minVotes)
  
jcitation_tidy$ifUnanimous[jcitation_tidy$ifUnanimous > 0] <- "majorityNonUnanimous"
jcitation_tidy$ifUnanimous[jcitation_tidy$ifUnanimous == 0] <- "majorityUnanimous"
  
jcitation_tidy <- jcitation_tidy %>%
  group_by(term, ifUnanimous) %>%
  summarize(number = n()) %>%
  ungroup() %>%
  spread(key = ifUnanimous, value = number) %>%
  mutate(number = majorityNonUnanimous + majorityUnanimous) %>%
  left_join(jcitation_ref, by = "term") %>%
  transmute(term, percentageTotal = number / totalNumber, percentageNonUnanimous = majorityNonUnanimous / nonUnanimous) %>%
  gather(percentageTotal, percentageNonUnanimous, key = type, value = percentage)

jcitation_tidy

ggplot(jcitation_tidy, aes(term, percentage, color = type)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1), labels = percent_format())

#4
#decision direction = 1 to all
jcitation_ref <- jcitation_all %>%
  distinct(caseIssuesId, .keep_all = TRUE) %>%
  group_by(term) %>%
  summarize(totalNumber = n())

jcitation_ref

jcitation_tidy <- jcitation_all %>%
  distinct(caseIssuesId, .keep_all = TRUE) %>%
  filter(decisionDirection == 1) %>%
  group_by(term) %>%
  summarize(number = n()) %>%
  left_join(jcitation_ref, by = "term") %>%
  mutate(percentage = number / totalNumber)

jcitation_tidy

ggplot(jcitation_tidy, aes(term, percentage)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1), labels = percent_format())

#5A
jcitation_chiefref1 <- jcitation_all %>%
  distinct(term, .keep_all = TRUE) %>%
  transmute(term, chief)

jcitation_chiefref1

jcitation_tidy <- jcitation_tidy %>%
  left_join(jcitation_chiefref1, by = "term")

jcitation_tidy

jcitation_chiefref2 <- jcitation_chiefref1 %>%
  group_by(chief) %>%
  summarize(chiefMean = mean(term), chiefSpan = max(term) - min(term)) %>%
  arrange(chiefMean) %>%
  mutate(chief = factor(chief, levels = chief))

jcitation_chiefref2

#solution 1
ggplot(jcitation_tidy, aes(term, percentage, color = chief)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1), labels = percent_format())

#solution 2
ggplot(jcitation_tidy, aes(term, percentage)) +
  geom_line() +
  geom_tile(data = jcitation_chiefref2, aes(chiefMean, 0.5, fill = chief, width = chiefSpan), alpha = 0.2) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format())
  
#reference
ggplot(jcitation_chiefref2, aes()) +
  geom_tile(aes(chiefMean, 0.5, fill = chief, width = chiefSpan), alpha = 0.2)

df <- data.frame(
  x = rep(c(2, 5, 7, 9, 12)),
  y = rep(c(1), 5),
  z = factor(1:5)
)

df
ggplot(df, aes()) +
  geom_tile(aes(x, y, fill = z), alpha = 0.2)

# Solution 3
# Overlap two line plots with facets by chief, while one of the dfs is without chief data therefore presented in all facets.
ggplot(jcitation_tidy, aes(term, percentage)) +
  geom_line(aes(term, percentage), mutate(jcitation_tidy, chief = NULL), alpha = 0.2) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  xlab("Term") +
  ylab("Percent of total decisions") +
  facet_wrap(~ chief)

```



at least one advanced challenge.
Code is well-documented (both self-documented and with additional comments as necessary).
Graphs and tables are properly labeled.
Use multiple commits to back up and show a progression in the work.
Analysis is clear and easy to follow, either because graphs are labeled clearly or you’ve written additional text to describe how you interpret the output.

Follow instructions on homework workflow to submit your homework. As part of the pull request, you're encouraged to reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc.

1. Lack of basic R knowledge make some operations in my mind couldn't be realized
I was confused by some of the basic operations in R. For example, I tried to make a for loop to change the value in df with certain criteria but failed finally.
Luckily, most problems solved after reading several more chapters from the textbook this week!

2. Lack of background knowledge add up some difficulty
No knowledge of legal system in the US makes me a little bit confused when reading definitions of the justice data. However, though can't understand fully the things described in the document, it is no problem to complete the work thanks to prof's demonstrative solution.

3. Technical difficulties
- Substitute values in cell.
A primal and brutal way to change the value arbitrarily.
Reference: http://stackoverflow.com/questions/5824173/replace-a-value-in-a-data-frame-based-on-a-conditional-if-statement-in-r

- Color shading
A little bit confusing when first looked at the tutorial. Tried them out and understood the arguments finally!
Reference: http://docs.ggplot2.org/current/geom_tile.html
